Omics-based Single Step Trait Prediction

### 2017-03-30 paper/manuscript.tex
Discuss SS-BLUP with SSBR, Add trait acronym table

% Introduction
\section{Introduction} 
% Rationale for predictions.
Genomic prediction, pioneered by \citeNP{Meuwissen2001}, has revolutionized
animal and plant breeding by accelerating the selection of promising candidates 
without phenotypes \cite{DeLosCampos2013,Garcia-Ruiz2016} and recently emerged
as an important tool in human diagnostics \cite{DeLosCampos2010,Vazquez2016}.
Its use is largely motivated by the high costs and oftentimes long duration of 
phenotyping, which can be reduced when covering all genotypes with a predictor 
while limiting phenotyping to a few genotypes with high precision
\cite{Kadam2016}.
The performance of all other genotypes is then forecast using model parameters
obtained from thorough (repeated) $k$-fold cross-validation routines applied to 
the subset of individuals with full information \cite{Technow2014}.
In animal breeding, genomic selection has increased the annual selection gain 
by $\approx$ 50-100\% for yield traits and by 300-400\% for traits with low 
heritability \cite{Garcia-Ruiz2016}.
In plant breeding, the relative excellence of a breeding strategy largely 
depends on the prediction accuracy, which may vary widely among traits and 
crops \cite{Longin2015}.

@@

% Workings and advantages of the single step procedure.
Independently from each other, \citeNP{Legarra2009} and
\citeNP{Christensen2010} developed a single step BLUP method that blends the 
numerator relationship matrix $\mathbf{A}$ with the genomic relationship matrix
$\mathbf{G}$ in a mutual matrix $\mathbf{H}$ to use all predictor information
simultaneously.
Issues with this H-BLUP approach were the compatability of $\mathbf{A}$ with
$\mathbf{G}$ \cite{Christensen2012} and their weighting, which is not
trait-independent \cite{Ashraf2016}.
\citeNP{Fernando2014} derived the equivalent single step marker-effect-model to 
the single step breeding value model of \citeNP{Legarra2009} and 
\citeNP{Christensen2010}, which has the advantage of not requiring weights for
$\mathbf{A}$ and $\mathbf{G}$ as well as the ability to model the structure of 
the imputation error.

@@

\section{Materials and Methods}
\Genetics2level{Maize hybrids}
The maize hybrid data set comprises 1,521 hybrids produced in 16 factorial
mating designs between 142 Dent and 103 Flint parent lines \cite{Westhues2017}. 
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse locations across Germany for six 
agronomically important traits.
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper \cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50 \cite{Ganal2011}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
using the Beagle software \cite{Browning2009}.
The final number of polymorphic marker loci was 7,013 for parental Dent and
6,212 for parental Flint lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 parental Dent and 43 parental Flint lines, that are parents of 685
hybrid progeny, through two-color hybridizations using a custom 2K microarray 
(GPL22267) \cite{Westhues2017}.
After applying established normalization procedures 
\cite{Smyth2003,Ritchie2007}, gene expression BLUEs were computed for 1,323 
transcripts in reference to established protocols 
\cite{Smyth2003,Ritchie2007,Frisch2010} using the R-package \textit{limma}
\cite{Ritchie2015a}.

\Genetics2level{Maize diversity panel}
The maize inbred line data set originally comprised 513 maize lines representing 
the global maize diversity and was reduced to the set of tropical and
subtropical lines ($n = 211$) given that it is the largest of the four
pre-classified subgroups \cite{Yang2014}.
All inbred lines were evaluated in five Chinese environments ranging from
$18$ to $30^{\circ}$N and from $102$ to $110^{\circ}$E \cite{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated for 17 traits of which 
six were used for this study.
MaizeSNP50 BeadChip-data were available for the entire set of lines from the
diversity panel.
Additionally, RNA sequencing was performed for 368 lines on young seedlings 15 
days after sowing.
By exploting identify by descent (IBD) between 49,728 SNPs on the BeadChip 
overlapping with the RNA-seq data, 556,809 high quality SNPs could be inferred 
for these lines \cite{Fu2013,Li2013}.
For the remaining 145 maize lines without RNA-seq data, high density markers
were inferred via projection of IBD regions onto the BeadChip, using this core 
set of markers \cite{Yang2014}.
The same SNP quality checks as for the parental inbred lines of the hybrids
were applied to the 211 inbred lines from the tropical/subtropical subset,
yielding 37,760 SNPs.
Gene expression data were normalized using a normal quantile transformation to
satisfy modelling assumptions \cite{Fu2013}.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes and was kept for further analyses \cite{Li2013}.

@@

\section*{Results}
\Genetics2level{Inbred lines}
At the level of the 149 genotypes for which both, genomic (G) and 
transcriptomic (T) information, were available, the use of T yielded higher
predictive abilities than the use of G for four out of six traits
(Fig. \ref{fig:InbredResults}A).
In terms of predictive ability, the absolute advantage of T over G was as large 
as 0.04 as observed for the traits DS, ED, GW and KW.
Except for KW, the predictive ability for predictions based on G were always
higher for the full set of inbred lines ($n = 211$) than for the reduced set of 
inbred lines ($n = 149$).
The combination of G and T (\textit{i.e.}, GT) in a single-step approach for the 
prediction of all 211 lines performed at least as well as G alone when T was 
superior over G for the reduced set of 149 lines.
In the case of GW and ED, the use of GT yielded a slight improvement in predictive 
ability over G.
Where T was a worse predictor for the reduced set of 149 lines than G,
predictive abilities of GT for the full set of 211 lines were also worse than 
those based on G alone.

@@

\Genetics2level{Hybrids}
The maize hybrid data set contains information on six agronomic traits and
three predictors, namely pedigree data (P), genomic information (G), 
transcriptomic information (T), and combinations thereof.
Pedigree records and genomic information were available for the 245 parent 
lines of all 1,521 hybrids and we denote these data as the 'Full' set whereas
transcriptomic data were available for 103 parent lines of 685 hybrids, which
we denote as 'Core'.
For dry matter yield (DMY), T was the best predictor in the core set of 685 
hybrids with a predictive ability that was 0.08 points higher than for genomic 
data, as the second best predictor for this trait 
(Fig. \ref{fig:HybridResults}).
In the core set, P had the highest predictive ability for starch ('STA') and 
dry matter content ('DMC').
For the other three traits, inside the core set G was the superior predictor.
In the full set of 1,521 hybrids, predictive abilities obtained with G rose 
markedly for all traits with the exception of 'FAT'.
The combination of genomic and transriptomic information (GT) in a single-step
model yielded a small improvement of G alone for the trait 'DMY'.
All traits, where T was not superior to G in the core set, did not benefit from 
the single-step approach when compared to genomic information alone.
Predictive abilities resulting from combinations of pedigree with genomic (PG)
information were better than P alone for all traits and PT improved upon P in
the core set for four traits.

@@

\Genetics2level{Coverage of the genetic space}
Single-step approaches are based on the use of one complete predictor that
covers all genotypes and one incomplete predictor that covers only a subset of
all available genotypes.
Here, we explore the impact of which data are covered by the incomplete 
predictor on the predictive abilities obtained via single-step approaches by
focusing exclusively on the 149 inbred lines from the maize diversity panel
that are covered by both, genomic and transcriptomic information.
First, we assmbled nine different core sets, which we defined as subsets
of genotypes that maximize the average genetic distance between genotypes of
the full set.
The nine core sets differ in that each covers a different number of genotypes,
ranging from only 10\% to to 90\% of all genotypes covered by the incomplete
predictor (Fig. \ref{fig:CoreSetPCA}).

@@

Predictive abilities obtained with the different core sets varied widely within
traits (Fig. \ref{fig:InbredResults}C).
For the traits CW and PH, which were better predicted via G than via T in the 
reduced set of lines, predictive abilities were higher when only a small number 
of genotypes was included in the core set.
The latter corresponds to a situation where a large number of genotypes is
being imputed and the majority of the information is derived from genomic
information. 
For most other traits, an initial increase in predictive abilities when moving
from a core set size of 0.1 to 0.5/0.6 was observed, followed by a drop around
0.6/0.7 and a subsequent hike up to a core set size of 0.9.
A principal component analysis depicts the coverage of the genetic space by
genotypes (Fig. \ref{fig:PopStructure}).
With a core set size of 50-60\% of the 149 genotypes, the congruence between
core set and complement was high whereas genotypes in the south-west direction
of the PCA plot were overrepresented compared to genotypes in the east 
direction at a core set size of 0.7.

As an alternative to exploring the impact of genotypes sampled based on the
maximization of genetic distance within the core set, we wanted to explore a
more extreme scenario where we select core sets based on population structure.
Hereto, we ran an admixture analysis on the same set of 149 inbred lines and
display the estimated ancestry coefficients for the genotypes assuming $K=3$
ancestral populations (Fig. \ref{fig:PopStructure}).

@@

Of the 149 genotypes, 17 belonged to more than 50\% to the first ancestral
population (A1), 109 belonged to more than 50\% to the second ancestral
population (A2) and 19 belonged to more than 50\% to the third ancestral
population while four genotypes could not be unequivocally assigned to either
ancestral population (mixed).
To adjust the size of the set of genotypes largely belonging to A2 to the other
two sets, we selected 19 genotypes from A2 using previously described core
sampling.
For single-step prediction, each of the three genotype sets were then declared
as having genomic and transcriptomic information whereas their complements were
covered by transcriptomic data only.
Genotypes from A1 and A3, which were at the far ends in the PCA plot (Fig.
\ref{fig:PopStructure}B), were estimated with less precision than genotypes
from the central cluster of genotypes belonging largely to the A2 population
(Fig. \ref{fig:PopStructure}A, Fig. \ref{fig:InbredResults}B).

A principal component analysis, carried out separately on the two parental
heterotic groups of the maize hybrids, showed that the entire genetic space in
each group was well covered by genotypes that had both, transcriptomic and
genomic data (Fig. \ref{fig:UHOH-PCA}).

@@

\section*{Discussion}
\Genetics2level{Addition of new predictors}
Pedigree records are the most ubiquitous predictor in animal and plant breeding
programs, because they are easy and inexpensive to collect.
Compared to pedigree information, which represent the expected relationship
among individuals, genomic information capture Mendelian sampling and thereby 
provide an improved proxy of the realized relationship among individuals.
Nevertheless, genomic data do not exhaustively capture physiological epistasis 
\cite{Jiang2015,Guo2016,Vazquez2016}, describing the interactions 
within and between different biological strata \cite{Sackton2016}.
Such interactions were found to be pervasive throughout the genomes of yeast 
\cite{Brem2005} and humans \cite{Brown2014} and have motivated studies on the 
utility of downstream 'omics' predictors for integrating such interactions.
Recently, encouraging results have been found for humans \cite{Vazquez2016}, 
maize inbred lines \cite{Guo2016} and maize hybrids \cite{Westhues2017}.
While single step prediction has become an established method when complete
pedigree but only incomplete genomic records are available, its application has
not yet been considered for the imputation of a complete predictor based on 
incomplete downstream 'omics' predictors such as transcriptomic, proteomic or 
metabolomic data.
Here, we considered single step approaches for a maize inbred line diversity
panel as well as for a collection of maize hybrids using pedigree, genomic and
transcriptomic data as predictors.

@@

\subsection{Maize hybrids}
Hybrid breeding --- with noteworthy commercial applications in pig, chicken and 
maize breeding --- is a particularly challenging application field for
prediction tools.
Here, $2n$ parent individuals from two genetically distinct heterotic groups are
crossed to each other, yielding $n^{2}$ potential hybrid progeny that would
require intensive field testing.
In medium-sized plant breeding programs, the advent of the doubled haploid (DH) 
technology \cite{Wedzony2009} now allows for an annual production of 1,000 
parent lines in each heterotic group, amounting to $10^{6}$ putative hybrid
progeny.
So far, its applications in hybrid populations are limited to pig breeding
\cite{Xiang2015,Xiang2016,Tusell2016}.
% Are there applications in chicken breeding? (Hints in Christensen et al.
% 2014, GSE).
For the hybrid data, the predictive abilities of genomic information
for the whole set of available parent lines rose considerably for several
agronomic traits.
Hence, the inclusion of trancsriptomic data could not further improve upon
this result.
Predictive abilities obtained when using pedigree data could be 
improved considerably when, depending on the agronomic trait, combined with 
either transcriptomic or genomic information.
Evaluations of core sets for the parental inbred lines of the maize hybrids
used throughout this study were outside the scope of this paper, but
results on the inbred lines have shown that homogeneous coverage of the
genetic space by the incomplete predictor can considerably leverage the
predictive ability in single step prediction compared to using only the
complete predictor.
In hybrid breeding, the probability of successfully selecting the best
observed genotypes based on the best predicted candidates is a strongly 
convex function of the predictive ability.
Hence, even the minor gains when using transcriptomic and genomic data in a
single step might justify additional investments in RNA-seq at least of a
subset of genotypes that represents the genetic space of the breeding 
material well.

\Genetics2level{Coverage of the genetic space}
Animal populations are so large that it does not truly matter, which animals are 
being genotyped, but in plant breeding the situation is different and we would 
be interested in determining which individuals should be complemented with 
information on another predictor.
If one can cover the genetic space sufficently with the incomplete predictor,

When using core sets of varying sizes, which were established with the
constraint of maximizing the average genetic distance among their members, we
noticed strong fluctuations in predictive abilities when moving from small core
sets to large core sets.
For traits that were better predicted by transcriptomic than by genomic
information in the core set, predictive abilities typically peaked around
core set size of 0.5/0.6, then dropped sharply at 0.7 and started to increase
again with a size of 0.9.
A principal component analysis showed that, at a core set size of 0.7, 
genotypes assigned to the A3 ancestral population (Fig. \ref{fig:PopStructure})
were largely elements of the core set whereas genotypes with a large share at 
the A1 ancestral population were predominantly excluded from the core set.
At a core set size of 90\% of the original size, the majority of the A1
genotypes were finally covered by the core set as well.
This result suggests a negative bias in the imputation of genotypes from the A1
population and could explain the drop in predictive ability at a core set size
of 0.7, which is in agreement with the reduced predictive abilities observed
when stripping A1 genotypes from their transcriptomic information (Fig.
\ref{fig:InbredResults}B).

\Genetics2level{Advantages of marker effect models}
% Describe the many hacks that have been developed to tackle problems with BLUP
% models.

>>>>

% Introduction
\section{Introduction} 
% Rationale for predictions.
Genomic prediction, pioneered by \citeNP{Meuwissen2001}, has revolutionized
animal and plant breeding by accelerating the selection of promising candidates 
without phenotypes \cite{DeLosCampos2013,Garcia-Ruiz2016} and recently emerged
as an important tool in human diagnostics \cite{DeLosCampos2010,Vazquez2016}.
Its use is largely motivated by the high costs and oftentimes long duration of 
phenotyping, which can be reduced when covering all genotypes with a predictor 
while limiting phenotyping to a few genotypes with high precision
\cite{Kadam2016}.
The performance of all other genotypes is then forecast using model parameters
obtained from thorough resampling routines applied to the subset of individuals 
with full information \cite{Hastie2009}.
In animal breeding, genomic selection has increased the annual selection gain 
by $\approx$ 50-100\% for yield traits and by 300-400\% for traits with low 
heritability \cite{Garcia-Ruiz2016}.
In plant breeding, the relative excellence of a breeding strategy largely 
depends on the prediction accuracy, which may vary widely among traits and 
crops \cite{Longin2015}.

@@

% Workings and advantages of the single step procedure.
Independently from each other, \citeNP{Legarra2009} and
\citeNP{Christensen2010} developed a single step BLUP method that blends the 
numerator relationship matrix $\mathbf{A}$ with the genomic relationship matrix
$\mathbf{G}$ in a mutual matrix $\mathbf{H}$ to use all predictor information
simultaneously.
Issues with this H-BLUP approach were the compatability of $\mathbf{A}$ with
$\mathbf{G}$ \cite{Christensen2012} and their weighting, which is not
trait-independent \cite{Vitezica2011,Ashraf2016}.
\citeNP{Fernando2014} derived the equivalent single step marker-effect-model to 
the single step breeding value model of \citeNP{Legarra2009} and 
\citeNP{Christensen2010}, which has the advantage of not requiring weights for
$\mathbf{A}$ and $\mathbf{G}$ as well as the ability to model the structure of 
the imputation error.

@@

\section{Materials and Methods}
\Genetics2level{Maize hybrids}
The maize hybrid data set comprises 1,521 hybrids produced in 16 factorial
mating designs between 142 Dent and 103 Flint parent lines \cite{Westhues2017}. 
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse locations across Germany for six 
agronomically important traits (Table \ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper \cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50 \cite{Ganal2011}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
using the Beagle software \cite{Browning2009}.
The final number of polymorphic marker loci was 7,013 for parental Dent and
6,212 for parental Flint lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 parental Dent and 43 parental Flint lines, that are parents of 685
hybrid progeny, through two-color hybridizations using a custom 2K microarray 
(GPL22267) \cite{Westhues2017}.
After applying established normalization procedures 
\cite{Smyth2003,Ritchie2007}, gene expression BLUEs were computed for 1,323 
transcripts in reference to established protocols 
\cite{Smyth2003,Ritchie2007,Frisch2010} using the R-package \textit{limma}
\cite{Ritchie2015a}.

\Genetics2level{Maize diversity panel}
The maize inbred line data set originally comprised 513 maize lines representing 
the global maize diversity and was reduced to the set of tropical and
subtropical lines ($n = 211$) given that it is the largest of the four
pre-classified subgroups \cite{Yang2014}.
All inbred lines were evaluated in five Chinese environments ranging from
$18$ to $30^{\circ}$N and from $102$ to $110^{\circ}$E \cite{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated for 17 traits of which 
six were used for this study (Table \ref{table:TraitKey}).
MaizeSNP50 BeadChip-data were available for the entire set of lines from the
diversity panel.
Additionally, RNA sequencing was performed for 368 lines on young seedlings 15 
days after sowing.
By exploting identify by descent (IBD) between 49,728 SNPs on the BeadChip 
overlapping with the RNA-seq data, 556,809 high quality SNPs could be inferred 
for these lines \cite{Fu2013,Li2013}.
For the remaining 145 maize lines without RNA-seq data, high density markers
were inferred via projection of IBD regions onto the BeadChip, using this core 
set of markers \cite{Yang2014}.
The same SNP quality checks as for the parental inbred lines of the hybrids
were applied to the 211 inbred lines from the tropical/subtropical subset,
yielding 37,760 SNPs.
Gene expression data were normalized using a normal quantile transformation to
satisfy modelling assumptions \cite{Fu2013}.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes and was kept for further analyses \cite{Li2013}.

% Please add the following required packages to your document preamble:
% \usepackage{booktabs}
\begin{table}[]
\centering
\caption{Acronyms for the trait names}
\label{table:TraitKey}
\begin{tabular}{@{}lll@{}}
\toprule
Trait              & Acronym & Data Set  \\ \midrule
Cob weight         & CW      & Diversity \\
Days to silking    & DS      & Diversity \\
Ear diameter       & ED      & Diversity \\
100-grain weight   & GW      & Diversity \\
Kernel width       & KW      & Diversity \\
Plant height       & PH      & Diversity \\
Dry matter yield   & DMY     & Hybrid    \\
Dry matter content & DMC     & Hybrid    \\
Fat                & FAT     & Hybrid    \\
Protein            & PRO     & Hybrid    \\
Starch             & STA     & Hybrid    \\
Sugar              & SUG     & Hybrid    \\ \bottomrule
\end{tabular}
\end{table}

@@

\section*{Results}
\Genetics2level{Inbred lines}
At the level of the 149 genotypes for which both, genomic (G) and 
transcriptomic (T) information, were available, the use of T yielded higher
predictive abilities than the use of G for four out of six traits
(Fig. \ref{fig:InbredResults}A).
In terms of predictive ability, the absolute advantage of T over G was as large 
as 0.04 as observed for the traits DS, ED, GW and KW.
Except for KW, the predictive ability for predictions based on G were always
higher for the full set of inbred lines ($n = 211$) than for the reduced set of 
inbred lines ($n = 149$).
The combination of G and T (\textit{i.e.}, GT) in a single-step approach for the 
prediction of all 211 lines, performed at least as well as G alone when T was 
superior over G for the reduced set of 149 lines.
In the case of GW and ED, the use of GT yielded a slight improvement in predictive 
ability over G.
Where T was a worse predictor for the reduced set of 149 lines than G,
predictive abilities of GT for the full set of 211 lines were also worse than 
those based on G alone.

@@

\Genetics2level{Hybrids}
The maize hybrid data set contains information on six agronomic traits and
three predictors, namely pedigree data (P), genomic information (G), 
transcriptomic information (T), and combinations thereof.
Pedigree records and genomic information were available for the 245 parent 
lines of all 1,521 hybrids and we denote these data as the 'Full' set whereas
transcriptomic data were available for 103 parent lines of 685 hybrids, which
we denote as 'Core'.
For DMY, T was the best predictor in the core set of 685 hybrids with a 
predictive ability that was 0.08 points higher than for genomic data, as the 
second best predictor for this trait (Fig. \ref{fig:HybridResults}).
In the core set, P had the highest predictive ability for STA and DMC.
For the other three traits, G was the superior predictor inside the core set.
In the full set of 1,521 hybrids, predictive abilities obtained with G rose 
markedly for all traits with the exception of FAT.
The combination of genomic and transriptomic information (GT) in a single-step
model yielded a small improvement over G alone for the traits DMY and PRO.
Except for PRO, compared to using only G, trait prediction did not benefit from 
the single step approach when T was not superior to G in the core set.
Predictive abilities resulting from combinations of pedigree with genomic (PG)
information were better than P alone for all traits and PT improved upon P in
the core set for four traits.

@@

\Genetics2level{Coverage of the genetic space}
Single-step approaches are based on the use of one complete predictor that
covers all genotypes and one incomplete predictor that covers only a subset of
all available genotypes.
Here, we explore the impact of which data are covered by the incomplete 
predictor on the predictive abilities obtained via single-step approaches by
focusing exclusively on the 149 inbred lines from the maize diversity panel
that are covered by both, genomic and transcriptomic information.
First, we assembled nine different core sets, which we defined as subsets
of genotypes that maximize the average genetic distance between genotypes of
the full set.
The nine core sets differ in that each covers a different number of genotypes,
ranging from only 10\% to 90\% of all genotypes covered by the incomplete
predictor (Fig. \ref{fig:CoreSetPCA}).

@@

Predictive abilities obtained with the different core sets varied widely within
traits (Fig. \ref{fig:InbredResults}B).
For the traits CW and PH, which were better predicted via G than via T in the 
reduced set of lines, predictive abilities were higher when only a small number 
of genotypes was included in the core set.
This corresponds to a situation where a large number of genotypes is being 
imputed and the majority of the information is derived from genomic information. 
For most other traits, an initial increase in predictive abilities when moving
from a core set size of 0.1 to 0.5/0.6 was observed, followed by a drop around
0.6/0.7 and a subsequent hike up to a core set size of 0.9.
A principal component analysis depicts the coverage of the genetic space by
genotypes (Fig. \ref{fig:PopStructure}).
With a core set size of 50-60\% of the 149 genotypes, the congruence between
core set and complement was high whereas genotypes in the south-west direction
of the PCA plot were overrepresented compared to genotypes in the east 
direction at a core set size of 0.7.

As an alternative to exploring the impact of genotypes sampled based on the
maximization of genetic distance within the core set, we wanted to explore a
more extreme scenario where we select core sets based on population structure.
Hereto, we ran an admixture analysis on the same set of 149 inbred lines 
assuming $K=3$ ancestral populations (Fig. \ref{fig:PopStructure}).

@@

Of the 149 genotypes, 17 belonged to more than 50\% to the first ancestral
population (A1), 109 belonged to more than 50\% to the second ancestral
population (A2) and 19 belonged to more than 50\% to the third ancestral
population while four genotypes could not be unequivocally assigned to either
ancestral population (mixed).
To reconcile the large size of the A2 set compared to A1 and A3, we selected 19 
genotypes from A2 using previously described core sampling.
For single-step prediction, each of the three genotype sets were then declared
as having genomic and transcriptomic information whereas their complements were
covered by transcriptomic data only.
Genotypes from A1 and A3, which were at the far ends in the PCA plot (Fig.
\ref{fig:PopStructure}B), were estimated with less precision than genotypes
from the central cluster of genotypes belonging largely to the A2 population
(Fig. \ref{fig:PopStructure}A, Fig. \ref{fig:InbredResults}C).

Regarding the hybrid material, a PCA carried out separately for the two groups
of parental inbred lines showed that the entire genetic space in each group 
was covered well by genotypes that had both, transcriptomic and genomic data 
(Fig. \ref{fig:UHOH-PCA}).

@@

\section*{Discussion}
\Genetics2level{Advantages of marker effect models}
Single step approaches in hybrid populations have so far been limited to pig
breeding \cite{Xiang2015,Xiang2016,Tusell2016}.
All of these studies used breeding value-based single step BLUP methods
(SS-BLUP) conceived by \citeNP{Legarra2009} and \citeNP{Christensen2010} with
extensions to hybrid populations \cite{Christensen2014,Christensen2015}.
An attactive property of SS-BLUP models is their computational advantage over 
marker-effect single step Bayesian regression models (SSBR) when the number of 
genotypes is smaller than the number of features of the incomplete predictor.
Nevertheless, SS-BLUP suffers from a variety of issues, which we will present
briefly:
First, the combination of the numerator with the genomic relationship matrix
requires commensurability of the two matrices, which necessitates some form of
scaling and weighting of $\mathbf{A}$ and $\mathbf{G}$
\cite{Christensen2012,Christensen2012a}.
Second, the fact that mostly recent individuals have genotypes whereas the 
majority of old individuals has only pedigree records introduces a bias that 
has to be accounted for \cite{Vitezica2011,Legarra2015,Garcia-Baccino2017}.
Third, the addition of new genotypes requires updates of $\mathbf{G}$ that
demand additional measures such as computing the inverse of $\mathbf{G}$ via
recursion
\cite{Misztal2014,Misztal2016,Misztal2016a,Fragomeni2015,Masuda2016,Pocrnic2016}.
Advantages of the SSBR algorithm, conceived by \citeNP{Fernando2014}, include
that i) it does not require commensurability of $\mathbf{A}$ and $\mathbf{G}$, 
ii) it models the bias incurred by the incomplete predictor explicitly as 
$\mathbf{\epsilon}$ and iii) its speed depends largely on the number of 
features.
To our best knowledge, this study presents the first implementation of the SSBR 
algorithm in a hybrid data set.

\Genetics2level{Addition of new predictors}
Pedigree records are the most ubiquitous predictor in animal and plant breeding
programs, because they are easy and inexpensive to collect.
Compared to pedigree information, which represent the expected relationship
among individuals, genomic information capture Mendelian sampling and thereby 
provide an improved proxy of the realized relationship among individuals.
Nevertheless, genomic data do not exhaustively capture physiological epistasis 
\cite{Jiang2015,Guo2016,Vazquez2016}, describing the interactions 
within and between different biological strata \cite{Sackton2016}.
Such interactions were found to be pervasive throughout the genomes of yeast 
\cite{Brem2005} and humans \cite{Brown2014} and have motivated studies on the 
utility of downstream 'omics' predictors for integrating such interactions.
Recently, encouraging results have been found for humans \cite{Vazquez2016}, 
maize inbred lines \cite{Guo2016} and maize hybrids \cite{Westhues2017}.
While single step prediction has become an established method when complete
pedigree but only incomplete genomic records are available, its application has
not yet been considered for the imputation of a complete predictor based on 
incomplete downstream 'omics' predictors such as transcriptomic, proteomic or 
metabolomic data.
Here, we considered single step approaches for a maize inbred line diversity
panel as well as for a collection of maize hybrids using pedigree, genomic and
transcriptomic data as predictors.

@@

\subsection{Maize hybrids}
Hybrid breeding --- with noteworthy commercial applications in pig, chicken and 
maize breeding --- is a particularly challenging application field for
prediction tools.
Here, $2n$ parent individuals from two genetically distinct heterotic groups are
crossed to each other, yielding $n^{2}$ potential hybrid progeny that would
require intensive field testing.
In medium-sized plant breeding programs, the advent of the doubled haploid (DH) 
technology \cite{Wedzony2009} now allows for an annual production of 1,000 
parent lines in each heterotic group, amounting to $10^{6}$ putative hybrid
progeny.
% Are there applications in chicken breeding? (Hints in Christensen et al.
% 2014, GSE).
For the hybrid data, the predictive abilities of genomic information
for the whole set of available parent lines rose considerably for several
agronomic traits.
Hence, the inclusion of trancsriptomic data could not further improve upon
this result.
Predictive abilities obtained when using pedigree data could be 
improved considerably when, depending on the agronomic trait, combined with 
either transcriptomic or genomic information.
Evaluations of core sets for the parental inbred lines of the maize hybrids
used throughout this study were outside the scope of this paper, but
results on the inbred lines have shown that homogeneous coverage of the
genetic space by the incomplete predictor can considerably leverage the
predictive ability in single step prediction compared to using only the
complete predictor.
In hybrid breeding, the probability of successfully selecting the best
observed genotypes based on the best predicted candidates is a strongly 
convex function of the predictive ability.
Hence, even the minor gains when using transcriptomic and genomic data in a
single step might justify additional investments in RNA-seq at least of a
subset of genotypes that represents the genetic space of the breeding 
material well.

\Genetics2level{Coverage of the genetic space}
Animal populations are so large that it does not truly matter, which animals are 
being genotyped once more than 10,000 animals have pedigree records 
\cite{Fragomeni2015,Lourenco2015,Masuda2016}, but in plant breeding programs,
pedigrees are typically considerably smaller.
Single step prediction offers the promise of leveraging the predictive ability
for all genotypes by borrowing the superior performance of an auxiliary
predictor on only a subset of genotypes, thereby reducing predictor costs.
Hence, we were interested in determining which individuals should be 
complemented with information on another predictor.
When using core sets of varying sizes, which were established with the
constraint of maximizing the average genetic distance among their members, we
noticed strong fluctuations in predictive abilities when moving from small core
sets to large core sets.
For traits that were better predicted by transcriptomic than by genomic
information in the core set, predictive abilities typically peaked around
core set size of 0.5/0.6, then dropped sharply at 0.7 and started to increase
again with a size of 0.9.
A principal component analysis showed that, at a core set size of 0.7, 
genotypes assigned to the A3 ancestral population (Fig. \ref{fig:PopStructure})
were largely elements of the core set whereas genotypes with a large share at 
the A1 ancestral population were predominantly excluded from the core set.
At a core set size of 90\% of the original size, the majority of the A1
genotypes were finally covered by the core set as well.
This result suggests a negative bias in the imputation of genotypes from the A1
population and could explain the drop in predictive ability at a core set size
of 0.7, which is in agreement with the reduced predictive abilities observed
when stripping A1 genotypes from their transcriptomic information (Fig.
\ref{fig:InbredResults}B).
